Using Metric Space Indexing for Complete and Efficient Record Linkage

### 2017-11-22 paper.tex
Update on Overleaf.

Record linkage is the process of identifying records that refer to the
same real-world entities, in situations where entity identifiers are
unavailable. Records are linked on the basis of similarity between
common attributes, with every pair of records being classified as a
link or non-link depending on the degree of similarity between them.
Record linkage is usually performed in a three-step process: first
groups of similar candidate records are identified using indexing; 
pairs within the same group are then compared in more detail, and
finally classified.
%
Even state-of-the-art indexing techniques, such as Locality Sensitive
Hashing, have potential drawbacks. They may fail to group together some
true matching records with high similarity. Conversely, they may group
records with low similarity, leading to high computational overhead.
%
We propose using metric space indexing to perform \emph{complete} record
linkage, which results in a parameter-free record linkage process
combining indexing, comparison and classification into a single step
delivering complete and efficient record linkage. Our experimental
evaluation on real-world datasets from several domains shows that
linkage using metric space indexing can yield better quality than
current indexing techniques, with similar execution cost, without the
need for domain knowledge or trial and error to configure the process.

@@

While indexing techniques facilitate efficient linkage of very large
datasets~\cite{Don15}, they generally achieve scalability at the cost
of reduced linkage quality, because potentially true matching record
pairs are removed in the indexing step, leading to a reduction in
recall of the final linkage result~\cite{Chr12}. A variety of indexing
techniques, discussed in more detail in the following section, have
been proposed, ranging from simple phonetic based blocking~\cite{Chr12}
and sorting of the datasets~\cite{Dra12} to locality sensitive hashing
based techniques~\cite{Kim10,Steorts2014}. Techniques for
unsupervised~\cite{Kej13,Ram15} and supervised~\cite{Bil06,Mic06}
learning of optimal blocking schemes have also been proposed.

@@

Traditional record linkage systems that perform indexing prior to
comparison and classification, as illustrated in left-hand side in
Fig.~\ref{fig-rl-process}, add a further practical complexity to the
process. Indexing, comparison and classification are often conducted
using algorithms and parameters selected based on technical and
domain expertise of the user of the record linkage system, followed
by a manual assessment (auditing) of the linkage
outcomes~\cite{Chr12}. If the quality of the resulting links is not
good enough for a certain application, the linkage process needs to
be repeated with different parameter settings and possibly also
alternative algorithms. This can result in a time-consuming and
labour-intensive iterative process. A major challenge is often the
heuristic nature of indexing, where the choice of a certain indexing
technique and its parameters (including which attributes to use in indexing) will determine the final outcome of a linkage.

@@

A different approach to indexing is the sorted neighbourhood
method~\cite{Mon96}, where the datasets to be linked are sorted
according to a \emph{sorting key} (usually a concatenation of the values
from several attributes). A sliding window is then moved over the
datasets and only records within the window are compared. Techniques
that adaptively shrink or expand the window size based on the
characteristics of the sorting key values have shown to improve both
linkage efficiency and quality~\cite{Dra12} over
%other sliding window
approaches that use fixed sizes windows.

@@

Metric space indexing (MSI) techniques provide indexing structures to
support the comparison of records in one set with those in another,
typically also offering similarity search operations. To create an MSI
data structure, it is necessary to define a distance measure between
records, with certain properties including the \emph{triangle
inequality}. Similarity search operations include
\emph{range-search(\textbf{q},d)}, where all records $\mathbf{r}$
in a database $\mathbf{D}$ within a distance $d$ of a query record
$\mathbf{q}$ are identified; \emph{nearest-neighbour(\textbf{q})},
returning the record $\mathbf{r} \in \mathbf{D}$ with smallest
distance to $\mathbf{q}$; and \emph{nearest-n(\textbf{q},n)}, returning
the $n$ closest records in $\mathbf{D}$ to $\mathbf{q}$.
\cite{Zezula2010} gives an overview of MSI techniques. Here we choose
one metric space indexing structure, the M-tree~\cite{paolociaccia2m},
and investigate its efficacy for record linkage.

@@

We address the following general record linkage problem: for two
datasets $\mathbf{D}_A$ and $\mathbf{D}_B$, we wish to find, for each
record in $\mathbf{D}_A$, all the records in $\mathbf{D}_B$ that match
it. We compare several algorithms for record linkage: traditional
blocking, one incomplete similarity search method, LSH-MinHash, and
one complete method, M-tree~\cite{paolociaccia2m}. We also use a
simple complete brute force technique as a baseline for comparison,
although this can only feasibly be applied to the smallest of our
datasets. All experiments have a number of parameters to configure
the search space and algorithm behaviour, including a
\emph{distance threshold}, $d$, specifying a maximum distance,
equivalent to a minimum similarity, for two records to be classified
as a link (i.e.\ referring to the same entity).

@@

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-authors}
\vspace{-6mm}
\caption{Blocking on `authors' field}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-title}
\vspace{-6mm}
\caption{Blocking on `title' field}
\end{subfigure} \vspace{5mm}

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-year}
\vspace{-6mm}
\caption{Blocking on `publication year' field}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-combined}
\vspace{-6mm}
\caption{Blocking using the union of all fields}
\end{subfigure} \vspace{5mm}

@@

Birth records were linked to death records, separately for the Skye and
Kilmarnock datasets, using M-tree and a range of LSH configurations. It
was not computationally feasible to run the brute force linker. Of the
incomplete linkers, LSH was selected as it gave slightly better results
for Cora. The shingle size was set to 2 for all the LSH experiments
reported, as this was found to give good results and LSH was not
especially sensitive to this parameter. Results for other shingle sizes
are omitted from this paper for brevity. A lower range of distance
thresholds was explored, based on domain knowledge of the datasets.

\Cref{demography-quality} plots (a) and (b) show the M-tree precision,
recall, and F-measure for various thresholds. In both datasets, the best
F-measure values are obtained with a low distance threshold of 2. Plots
(c) and (d) compare the F-measure curves for M-tree with those obtained
from a range of LSH configurations. The best F-measure value for M-tree
is higher than that of any of the LSH configurations, for both datasets.
This demonstrates both the competitiveness of M-tree with respect to
linkage quality, and its important characteristic of being
parameter-free---the linkage quality is obtained without the need to
tuning for the dataset.

\Cref{comparison-of-results-demography-skye} and
\Cref{comparison-of-results-demography-kili} show greater detail for
selected linkers. In both cases the F-measure achieved is better for
M-tree than any of the LSH linkers. The better linkage quality achieved
by M-tree is largely due to recall for M-tree being much higher than for
any of the LSH configurations. In most cases, LSH out-performs M-tree in
terms of precision. More significantly, LSH linkage quality is heavily
dependent on the configuration parameters. For plausible settings for
the number of bands and band size, F-measure varies from 0.01 (extremely
poor) to 0.47 (relatively good) for Skye and from 0.03 to 0.49 for
Kilmarnock. In both cases \emph{LSH-10-2} performs best, but since this
is data-dependent there is no guarantee that these parameters would work
well with another dataset.

@@

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotFs-skye-f}
\vspace{-6mm}
\caption{F1 Measure on Skye dataset \\ with M-tree and all LSH configurations}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotFs-kilmarnock-f}
\vspace{-6mm}
\caption{F1 Measure on Kilmarnock dataset \\ with M-tree and all LSH configurations}
\end{subfigure}

\caption{Linkage results on the demographic datasets}
\label{demography-quality}
\end{figure}

>>>>

Record linkage is the process of identifying records that refer to the
same real-world entities, in situations where entity identifiers are
unavailable. Records are linked on the basis of similarity between
common attributes, with every pair being classified as a link or
non-link depending on their degree of similarity. Record linkage is
usually performed in a three-step process: first groups of similar
candidate records are identified using indexing; pairs within the same
group are then compared in more detail, and finally classified.
%
Even state-of-the-art indexing techniques, such as Locality Sensitive
Hashing, have potential drawbacks. They may fail to group together some
true matching records with high similarity. Conversely, they may group
records with low similarity, leading to high computational overhead.
%
We propose using metric space indexing to perform \emph{complete} record
linkage, which results in a parameter-free record linkage process
combining indexing, comparison and classification into a single step
delivering complete and efficient record linkage. Our experimental
evaluation on real-world datasets from several domains shows that
linkage using metric space indexing can yield better quality than
current indexing techniques, with similar execution cost, without the
need for domain knowledge or trial and error to configure the process.

@@

While indexing techniques facilitate efficient linkage of very large
datasets~\cite{Don15}, they generally achieve scalability at the cost
of reduced linkage quality, because potentially true matching record
pairs are removed in the indexing step, leading to a reduction in
recall~\cite{Chr12}. A variety of indexing
techniques, discussed in more detail in the following section, have
been proposed, ranging from simple phonetic based blocking~\cite{Chr12}
and sorting of the datasets~\cite{Dra12} to locality sensitive hashing
based techniques~\cite{Kim10,Steorts2014}. Techniques for
unsupervised~\cite{Kej13,Ram15} and supervised~\cite{Bil06,Mic06}
learning of optimal blocking schemes have also been proposed.

@@

Traditional record linkage systems that perform indexing prior to
comparison and classification, as illustrated on the left in
\Cref{fig-rl-process}, add a further practical complexity to the
process. Indexing, comparison and classification are often conducted
using algorithms and parameters selected based on technical and
domain expertise of the user of the record linkage system, followed
by a manual assessment (auditing) of the linkage
outcomes~\cite{Chr12}. If the quality of the resulting links is not
good enough for a certain application, the linkage process needs to
be repeated with different parameter settings and possibly also
alternative algorithms. This can result in a time-consuming and
labour-intensive iterative process. A major challenge is often the
heuristic nature of indexing, where the choice of a certain indexing
technique and its parameters (including which attributes to use in indexing) will determine the final outcome of a linkage.

@@

A different approach to indexing is the sorted neighbourhood
method~\cite{Mon96}, where the datasets to be linked are sorted
according to a \emph{sorting key} (usually a concatenation of the values
from several attributes). A sliding window is then moved over the
datasets and only records within the window are compared. Techniques
that adaptively shrink or expand the window size based on the
characteristics of the sorting key values have been shown to improve both
linkage efficiency and quality~\cite{Dra12} over
%other sliding window
approaches that use fixed size windows.

@@

Metric space indexing (MSI) techniques provide indexing structures to
support the comparison of records in one set with those in another,
typically also offering similarity search operations. To create an MSI
data structure, it is necessary to define a distance measure between
records, with certain properties including the \emph{triangle
inequality}. Similarity search operations include
\emph{range-search(\textbf{q},d)}, where all records $\mathbf{r}$
in a database $\mathbf{D}$ within a distance $d$ of a query record
$\mathbf{q}$ are identified; \emph{nearest-neighbour(\textbf{q})},
returning the record $\mathbf{r} \in \mathbf{D}$ with smallest
distance to $\mathbf{q}$; and \emph{nearest-n(\textbf{q},n)}, returning
the $n$ closest records in $\mathbf{D}$ to $\mathbf{q}$.
\cite{Zezula2010} gives an overview of MSI techniques. Here we choose
one metric space indexing structure, the M-tree~\cite{Ciaccia97indexingmetric},
and investigate its efficacy for record linkage.

@@

We address the following general record linkage problem: for two
datasets $\mathbf{D}_A$ and $\mathbf{D}_B$, we wish to find, for each
record in $\mathbf{D}_A$, all the records in $\mathbf{D}_B$ that match
it. We compare several algorithms for record linkage: traditional
blocking, one incomplete similarity search method, LSH-MinHash, and
one complete method, M-tree. We also use a
simple complete brute force technique as a baseline for comparison,
although this can only feasibly be applied to the smallest of our
datasets. All experiments have a number of parameters to configure
the search space and algorithm behaviour, including a
\emph{distance threshold}, $d$, specifying a maximum distance,
equivalent to a minimum similarity, for two records to be classified
as a link (i.e.\ referring to the same entity).

@@

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-authors}
\vspace{-6mm}
\caption{Blocking on `authors' attr.}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-title}
\vspace{-6mm}
\caption{Blocking on `title' attr.}
\end{subfigure} \vspace{5mm}

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-year}
\vspace{-6mm}
\caption{Blocking on `publication year' attr.}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-combined}
\vspace{-6mm}
\caption{Blocking using the union of all attrs.}
\end{subfigure} \vspace{5mm}

@@

Birth records were linked to death records, separately for the Skye and
Kilmarnock datasets, using M-tree and a range of LSH configurations. It
was not computationally feasible to run the brute force linker. Of the
incomplete linkers, LSH was selected as it gave slightly better results
for Cora. The shingle size was set to $l_{ss}=2$ for all the LSH
experiments reported, as this was found to give good results and LSH
was not especially sensitive to this parameter. Results for other
shingle sizes are omitted from this paper for brevity. A lower range
of distance thresholds was explored, based on domain knowledge of the
datasets.

\Cref{demography-quality} plots (a) and (b) show the M-tree precision,
recall, and F-measure for various thresholds. In both datasets, the best
F-measure values are obtained with a low distance threshold of $d=2$.
Plots (c) and (d) compare the F-measure curves for M-tree with those
obtained from a range of LSH configurations. The best F-measure value
for M-tree is higher than that of any of the LSH configurations, for
both datasets. This demonstrates both the competitiveness of M-tree
with respect to linkage quality, and its important characteristic of
being parameter-free---the linkage quality is obtained without the
need to tuning for the dataset.

\Cref{comparison-of-results-demography-skye} and
\Cref{comparison-of-results-demography-kili} show greater detail for
selected linkers. In both cases the F-measure achieved is better for
M-tree than any of the LSH linkers. The better linkage quality achieved
by M-tree is largely due to recall for M-tree being much higher than for
any of the LSH configurations. In most cases, LSH out-performs M-tree in
terms of precision. More significantly, LSH linkage quality is heavily
dependent on the configuration parameters. For plausible settings for
the number of bands and band size, F-measure varies from $0.01$
(extremely poor) to $0.47$ (relatively good) for Skye and from $0.03$
to $0.49$ for Kilmarnock. In both cases \emph{LSH-10-2} performs best,
but since this is data-dependent there is no guarantee that these
parameters would work well with another dataset.

@@

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotFs-skye-f}
\vspace{-6mm}
\caption{F1 Measure on Skye dataset \\ with M-tree and all LSH configurations}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotFs-kilmarnock-f}
\vspace{-6mm}
\caption{F1 Measure on Kilmarnock dataset \\ with M-tree and all LSH configurations}
\end{subfigure}
\caption{Linkage results on the demographic datasets}
\label{demography-quality}
\end{figure}
